---
title: "Final project"
author: "Mavis Fosuaa Boateng"
date: "2024-07-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load packages 
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, gapminder,
               here,visdat, 
               inspectdf,
               countrycode,   bslib, bsicons, shiny, 
               rnaturalearth, plotly, htmltools, 
               reactable,sf,ggplot2,sf, 
               rgeoboundaries
               )

```

# Data Importation

Raw dataset is about:

1.The total number of cases of malaria reported within a year. Malaria cases reported. (*C:/Users/mboateng21/Downloads/malaria_number_of_cases_reported.csv*)


```{r}
TB_old_cases <- read.csv(here("C:/Users/mboateng21/Downloads/TB/sh_tbs_incd.csv"))

TB_new_cases<-read.csv(here("C:/Users/mboateng21/Downloads/TB/all_forms_of_tb_incidence_estimated.csv"))

TB_deaths<-read.csv(here("C:/Users/mboateng21/Downloads/TB/all_forms_of_tb_deaths_per_100000_estimated.csv"))

TB_treated<-read.csv(here("C:/Users/mboateng21/Downloads/TB/sh_tbs_cure_zs.csv"))
```



```{r}
# Show the first 6 rows
head(TB_old_cases)
str(TB_old_cases)

head(TB_new_cases)
str(TB_new_cases)

head(TB_deaths)
str(TB_deaths)

head(TB_treated)
str(TB_treated)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

```{r}
# Create visdat plot
vis_dat (TB_cases)
```

```{r}
inspect_cat(TB_old_cases) %>%
 show_plot()

inspect_cat(TB_new_cases) %>%
 show_plot()
```

# Data Cleaning

## Pivot to Long Format

Converting the datasets from wide to long format will make merging easier.

#Data cleaning for old TB Cases
```{r long-format-total}
# Filter columns to only include the years 2010 to 2020
TB_cases_filtered  <- TB_old_cases %>%
  select(country, `X2010`: `X2020`)

# Converting from wide to long format
TB_cases_long <-  TB_cases_filtered%>%
  pivot_longer(
    cols = -country,       # Exclude the 'country' column
    names_to = "year",     # Name of the new 'year' column
    values_to = "cases" # Name of the new 'total_cases' column
  )

# View the long format data
print(TB_cases_long)



# Remove 'X' from the 'year' column
TB_cases_long$year <- gsub('X', '', TB_cases_long$year)

# Convert 'year' column to numeric
TB_cases_long$year <- as.numeric(TB_cases_long$year)

# View the updated dataframe
print(TB_cases_long)

# Remove the decimal point and convert back to numeric
TB_cases_long$cases <- as.numeric(as.integer(TB_cases_long$cases))

# Print the updated dataframe
print(TB_cases_long)

```




## Converting the TB_new_cases dataset from wide to long format and other data cleaning


```{r long-format-total}
# Filter columns to only include the years 2010 to 2020
TB_new_cases_filtered  <- TB_new_cases %>%
  select(country, `X2010`: `X2020`)

# Converting from wide to long format
TB_new_long <-  TB_new_cases_filtered%>%
  pivot_longer(
    cols = -country,       # Exclude the 'country' column
    names_to = "year",     # Name of the new 'year' column
    values_to = "new_cases" # Name of the new 'total_cases' column
  )

# View the long format data
print(TB_new_long)


# Remove 'X' from the 'year' column
TB_new_long$year <- gsub('X', '', TB_new_long$year)

# Convert 'year' column to numeric
TB_new_long$year <- as.numeric(TB_new_long$year)
# View the updated dataframe
print(TB_new_long)



# Convert data types from chr to num

TB_new_long <- TB_new_long %>%
  mutate(
    # Convert year to numeric
    year = as.numeric(year),
    # Remove 'k' and 'M' and convert to numeric
    new_cases= case_when(
      str_detect(new_cases, 'k$') ~ as.numeric(str_remove(new_cases, 'k')) * 1000,
      str_detect(new_cases, 'M$') ~ as.numeric(str_remove(new_cases, 'M')) * 1000000,
      TRUE ~ as.numeric(new_cases)
    ))
```





## Data cleaning for number of TB deaths.Converting the dataset from wide to long format.


```{r long-format-total}
# Filter columns to only include the years 2010 to 2020
TB_deaths_filtered  <- TB_deaths %>%
  select(country, `X2010`: `X2020`)

# Converting from wide to long format
TB_mortality <-  TB_deaths_filtered%>%
  pivot_longer(
    cols = -country,       # Exclude the 'country' column
    names_to = "year",     # Name of the new 'year' column
    values_to = "deaths" # Name of the new 'total_cases' column
  )

# View the long format data
print(TB_mortality)


# Remove 'X' from the 'year' column
TB_mortality$year <- gsub('X', '', TB_mortality$year)

# Convert 'year' column to numeric
TB_mortality$year <- as.numeric(TB_mortality$year)

# View the updated dataframe
print(TB_mortality)

# Remove the decimal point and convert back to numeric
TB_mortality$deaths <- as.numeric(as.integer(TB_mortality$deaths))

# Print the updated dataframe
print(TB_mortality)

```


## Data cleaning for number of TB cases treatment.Converting the dataset from wide to long format.


```{r long-format-total}
# Filter columns to only include the years 2010 to 2020
TB_treated_filtered  <- TB_treated %>%
  select(country, `X2010`: `X2020`)

# Converting from wide to long format
TB_cases_treated <-  TB_treated_filtered%>%
  pivot_longer(
    cols = -country,       # Exclude the 'country' column
    names_to = "year",     # Name of the new 'year' column
    values_to = "treated" # Name of the new 'total_cases' column
  )

# View the long format data
print(TB_cases_treated)


# Remove 'X' from the 'year' column
TB_cases_treated$year <- gsub('X', '', TB_cases_treated$year)

# Convert 'year' column to numeric
TB_cases_treated$year <- as.numeric(TB_cases_treated$year)

# View the updated dataframe
print(TB_cases_treated)


```
## Joining Datasets

Join the datasets using the common keys, `country` and `year`.

```{r join-datasets}

# Perform the left join
TB_cases_combined <- left_join(TB_cases_long,TB_new_long,
                              by = c("country", "year"))

TB_combined <- left_join(TB_cases_combined,TB_mortality,
                           by = c("country", "year"))

TB_data <- left_join(TB_combined, TB_cases_treated, 
                     by = c("country", "year"))

# View the combined dataset
print(TB_data)


#Remove unwanted column and rename

#library(dplyr)
#TB_data <- TB_data %>%
 # select( -estimated.y)

#TB_data <- TB_data %>%
  #rename(estimated = estimated.x)




```



```{r}
TB_dataset <- TB_data %>%
  mutate(total_cases = cases+ new_cases)

# Print the updated data frame
print("Updated Data Frame with new_column:")
print(TB_dataset)
```





# Line graphs showing the confirmed cases of TB cases among three selected countries 
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)


# Correct number of repetitions for the 'country' column
countries <- rep(c("Zimbabwe", "South Africa", "India"), each = 11)

# Correct number of years
years <- rep(2010:2020, times = 3)


# Create the combined plot
plot <- ggplot(TB_dataset, aes(x = year, y = total_cases, color = country)) +
  geom_line(size = 1) +                   # Line size
  geom_point(size = 3) +                  # Point size
  labs(
    title = "TB Cases in Zimbabwe, South Africa, and India",
    x = "Year",
    y = "TB Cases"
  ) +
  theme_minimal(base_size = 15) +         # Base font size for theme
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 20),  # Center the title
    axis.title.x = element_text(size = 14, face = "bold"),             # X-axis title
    axis.title.y = element_text(size = 14, face = "bold"),             # Y-axis title
    axis.text.x = element_text(angle = 45, hjust = 1),                 # X-axis text
    panel.grid.major = element_line(color = "gray", size = 0.5),       # Major grid lines
    panel.grid.minor = element_line(color = "lightgray", size = 0.25)  # Minor grid lines
  ) +
  scale_color_brewer(palette = "Set1")    # Color palette for the lines

# Print the combined plot
print(plot)
```


```{r}
# Correct number of repetitions for the 'country' column
countries <- rep(c("Zimbabwe", "South Africa", "India"), each = 11)

# Correct number of years
years <- rep(2010:2020, times = 3)


# List of countries to plot
country_list <- c("Zimbabwe", "South Africa", "India")

# Loop through each country and create the plot
for (i in country_list) {
  # Subset data for the current country
  country_data <- subset(TB_dataset, country == i)
  
  # Create the plot
  plot <- ggplot(country_data, aes(x = year, y = total_cases)) +
    geom_line(color = "steelblue", size = 1) +          # Line color and size
    geom_point(color = "darkred", size = 3) +           # Points color and size
    labs(
      title = paste("TB Cases in", i),
      x = "Year",
      y = "total_cases"
    ) +
    theme_minimal(base_size = 15) +                     # Base font size for theme
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 20),  # Center the title
      axis.title.x = element_text(size = 14, face = "bold"),             # X-axis title
      axis.title.y = element_text(size = 14, face = "bold"),             # Y-axis title
      axis.text.x = element_text(angle = 45, hjust = 1),                 # X-axis text
      panel.grid.major = element_line(color = "gray", size = 0.5),       # Major grid lines
      panel.grid.minor = element_line(color = "lightgray", size = 0.25)  # Minor grid lines
    )+
  scale_color_brewer(palette = "Set1")    # Color palette for the lines
  
  # Print the plot
  print(plot)
}

```




## Standardizing Country Names

Converting Country Names to ISO Codes: You can use the `countrycode()` function to map your dataset's country names to their corresponding ISO codes.

For example here I add ISO codes to each dataset for a consistent joining key.

```{r add-iso}
TB_dataset <- TB_dataset %>%
  mutate(country_iso = countrycode(country, 
                                   "country.name", 
                                   "iso3c"))
```

# Adding population data


```{r}
pop <- tidyr::population

pop
 
```





Now, do the same for the `pop` data frame:

```{r}
# Add 3- letter ISO code to pop
pop_iso <- pop %>%
mutate(country_iso = countrycode(country, 
                                   "country.name", 
                                   "iso3c"))
```


## Joining with Population Data


```{r population-join}
# Ensure the 'year' column in both datasets is of the same type
pop_iso <- pop_iso %>%
  mutate(year = as.character(year))  # Convert 'year' to character




# Ensure the 'year' column in hiv_data_combined is also of type character
TB_dataset <- TB_dataset %>%
  mutate(year = as.character(year))  # Convert 'year' to character

# Join on ISO code and year
final_dataset <-
  left_join(pop_iso,TB_dataset,
by =c("country_iso","year") )

final_dataset <- final_dataset%>% rename(country = country.x)  
  
```


 
Remove unwanted columns
```{r}
#Remove unwanted and duplicate columns named
library(dplyr)

# Assuming final_dataset is your dataframe and you want to remove the columns "country.y", "cases.x", and "cases.y"
data_complete <- final_dataset %>%
  select(-country.y, -cases.x, -cases.y)

# Print the updated dataframe
print(data_complete)


```

```{r}
#Print the first few rows of the transformed data
print(head(data_complete, 20))

# Check for NA values in the total_cases column
na_count <- sum(is.na(data_complete$malaria_cases_numeric))
print(paste("Number of NA values in malaria_cases_numeric:", na_count))

```










# Mapping country borders with `{rnaturalearth}`

• Let's draw a world map with country borders.

• `{rnaturalearth}` can map all the countries in the world, among others.

• Use `ne_countries()` with the `returnclass = "sf"` argument.


```{r}
# Load the map data
countries <- ne_countries(scale="medium", returnclass = "sf")
ggplot(data = countries) +
geom_sf()


# Filter the data for the latest year available
latest_year <- final_dataset %>%
  filter(!is.na(year)) %>%
  summarise(latest_year = max(as.numeric(year), na.rm = TRUE)) %>%
  pull(latest_year)

latest_data <- final_dataset %>%
  filter(year == latest_year)
```


```{r}

world_data <- countries %>%
  left_join(latest_data, by = c("iso_a3" = "country_iso"))

# Create the choropleth map
ggplot(data = world_data) +
  geom_sf(aes(fill =total_cases ), color = "grey", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", trans = "log10", na.value = "darkblue", 
                       name = "TB_cases") +
  theme_minimal() +
  labs(
    title = paste("Global Tuberculosis prevalence in", latest_year),
    subtitle = "Choropleth map showing Tuberculosis cases per 100k population",
    caption = "Source: Tuberculosis data and rnaturalearth dataset"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

```



